---
title: "Is the New ABS System Useful/Needed?"
subtitle: "Umpire Accuracy in the 2025 World Series"
author: "David Zuhlsdorf"
format: revealjs
editor: visual
---

## Data Introduction

- This project uses pitch-level Statcast data from the **2025 World Series**.

- The goal is to compare real umpire ball/strike calls to an estimated automated strike zone.

```{r}
library(tidyverse)
library(readxl)
library(plotly)
library(here)
library(viridis)

data <- read_excel(here::here("data/2025WS_data.xlsx"))
```

---

## Question(s) of Interest

**Main question:**  
Would an ABS-style strike zone have been useful in the 2025 World Series?

More specifically:

- Where did umpires miss ball/strike calls?
- Were missed calls concentrated near the edges of the zone?
- Did accuracy change by game situation or count?


```{r}
data$plate_x <- as.numeric(data$plate_x)
data$plate_z <- as.numeric(data$plate_z)

clean_data <- data %>%
  filter(des %in% c("Ball", "Called Strike")) %>%
  drop_na(plate_x, plate_z, sz_top, sz_bot) %>%
  mutate(
    ump_call = if_else(des == "Called Strike", "Strike", "Ball"),
    abs_call = if_else(
      plate_x >= -0.83 & plate_x <= 0.83 &
        plate_z >= sz_bot & plate_z <= sz_top,
      "Strike", "Ball"
    ),
    correct_call = if_else(ump_call == abs_call, "Yes", "No"),
    z_norm = (plate_z - sz_bot) / (sz_top - sz_bot),
    count = paste(balls, strikes, sep = "-"),
    late_inning = ifelse(inning >= 7, "Yes", "No"),
    run_diff = abs(home_score - away_score),
    close_game = ifelse(run_diff <= 4, "Yes", "No"),
    high_lev = ifelse(
      late_inning == "Yes" & close_game == "Yes",
      "High leverage situation",
      "Not high leverage situation"
    )
  )
```



## Correct and Incorrect Calls

```{r}
ggplot(clean_data, aes(x = plate_x, y = z_norm, color = correct_call)) +
  geom_point(position = "jitter", alpha = 0.6) +
  geom_rect(
    xmin = -0.83, xmax = 0.83,
    ymin = 0, ymax = 1,
    fill = NA, color = "black", linewidth = 1
  ) +
  theme_minimal(base_size = 14) +
  scale_color_viridis_d() +
  labs(
    title = "Correct and Incorrect Ball/Strike Calls",
    x = "Horizontal Pitch Location",
    y = "Normalized Strike Zone Height",
    color = "Correct Call"
  ) +
  facet_wrap(~high_lev)
```



## Accuracy by Game

```{r}
# overall accuracy bar chart 
accuracy_data <- clean_data %>%
  group_by(game_date) %>%
  count(correct_call) %>%
  mutate(game_date = as.character(game_date),
    accuracy = (n / sum(n))*100) %>%
  filter(correct_call == "Yes")

barplot<- ggplot(accuracy_data,aes(x = game_date, y = accuracy, fill = accuracy, label = accuracy)) +
  geom_col() +
  theme_minimal() +
  scale_fill_viridis_c() +
  geom_hline(aes(yintercept = 94.15,
             linetype = "2025 Average Accuracy:94.15%")) +
  labs(
    title = "Umpire Accuracy by World Series Game",
    x = "Game Date",
    y = "Accuracy (%)",
    linetype = ""
  )

ggplotly(barplot,tooltip = "label")
  
```



## Missed Call Density

```{r}
missed_calls <- clean_data %>%
  filter(correct_call == "No")

ggplot(missed_calls, aes(x = plate_x, y = z_norm)) +
  stat_density_2d(
    aes(fill = after_stat(density)),
    geom = "raster",
    contour = FALSE
  ) +
  geom_rect(
    xmin = -0.83, xmax = 0.83,
    ymin = 0, ymax = 1,
    fill = NA, color = "white", linewidth = 1
  ) +
  scale_fill_viridis_c() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Density of Missed Ball/Strike Calls",
    x = "Horizontal Pitch Location",
    y = "Normalized Strike Zone Height",
    fill = "Density"
  )
```



## Accuracy by Count

```{r}
accuracy_by_count <- clean_data %>%
  group_by(count) %>%
  summarise(
    total_pitches = n(),
    correct_calls = sum(correct_call == "Yes"),
    accuracy = (correct_calls / total_pitches)*100
  )
count_plot <- ggplot(accuracy_by_count, aes(x = reorder(count, accuracy), y = accuracy, label = accuracy)) +
  geom_segment(aes(xend = count, y = 0, yend = accuracy), color = "steelblue") +
  geom_point(color = "red", size = 3) +
  coord_flip() +
  theme_minimal(base_size = 14) + 
  labs(
    title = "Umpire Accuracy by Count",
    x = "Count",
    y = "Accuracy (%)"
  ) 

ggplotly(count_plot, tooltip = "label")
```



## Main Takeaways

- Umpires were mostly accurate overall.
- Missed calls were most common near the edges of the strike zone and in non 2 strike counts
- ABS would likely be most useful on borderline pitches.
- The biggest value of ABS may be reducing mistakes in high-leverage situations.

## Limitations and Future Work

- This is an approximation of ABS, not official ABS data.
- One World Series is a small sample.
- This project does not fully simulate how game outcomes would change.
- Future work could include full-season data or run expectancy.
